---
title: "A Simple Interface To View Stock Prices"
author: "Matthew Halbasch"
date: "April 2, 2019"
output: ioslides_presentation
---

```{r setup, include=FALSE, message = FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(reshape2)
library(ggsci)
```

## Functionality

We built a simple [web app](https://mghalbasch.shinyapps.io/Data_Products_Final_Project/) that allows the user to input stock symbols and see how the price of that stock has changed over time.
The basic functionality includes:

- A slider to choose what years to display
- An input box to type in the stock symbol
    * Web app checks that the symbol matches a stock before displaying it
- A button to reset the page to start over if desired


## Usage

- To get started, just type a stock symbol into the box and hit "Submit"!

- A graph will be displayed with the stock's price over the years chosen

- If you add another stock symbol, the graph will be updated to compare the two.

- To change the range of dates, just move the slider for the min and max years.


## Data Source

The data for this web app comes from Yahoo Finance, and is loaded in via the
`quantmod` package.

For Google's stock (symbol GOOGL), the server would read in the data like this:

```{r echo = TRUE, message = FALSE}
library(quantmod)
df <- getSymbols("GOOG", auto.assign = FALSE)
```

We also clean the data briefly using the code:

```{r echo = TRUE}
df <- data.frame(df) %>% rownames_to_column("Date") %>%
        select(Date, GOOGL = ends_with("Close"))
```

## Example Plot

Finally we have an example of the kind of plot the web app produces.
This is made with the `ggplot2` package.

```{r}
df %>% melt(id.vars = "Date") %>% 
      ggplot(aes(x = as.Date(Date), y = value, color = variable)) +
      geom_line() +
      xlab("Date") +
      ylab("Closing Stock Price") +
      labs(title = "Stock Prices Over Time") +
      theme_minimal() +
      scale_color_nejm() +
      theme(legend.title = element_blank())
```


