---
title: "How Americans Have Viewed Their Most Important Problems Since 1960"
author: "Matthew Halbasch"
date: "April 2, 2019"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(plotly)
library(tidyverse)
```

## Gallup Survey Responses

- Gallup has asked American's to identify their most important problem dating
back to the 1950s.
- We will take a look at how Americans have responded, from 1960 to today.
- Data for this project come from the `politicaldata` package by G. Elliot Morris (see his blog, [The Crosstab](https://www.thecrosstab.com/project/politicaldata-package/), for more information)



## Cleaning the Data

We prepare the data by selecting responses from 1960 to today, and finding the rank of each response in a given year:

```{r echo=TRUE}
library(politicaldata)

pol_data <- cap_get_mip(min_year = 1960)
pol_data %>% group_by(year) %>% 
  mutate(rank = order(order(percent, decreasing = TRUE)),
         title_text = gsub("^General ", "", title_text)) %>%
  select(percent, year, title_text, rank) ->
  cleaned_data
```


## Animated Plot of the Data

```{r warning = FALSE, fig.height = 4, fig.width = 6, message = FALSE}
cleaned_data %>% plot_ly(x = ~title_text, y = ~percent, type = "scatter", 
                         frame = ~year, size = ~(1/rank)) %>%
  layout(showlegend = FALSE)
```

## Conclusion

- Americans frequently view domestic macroeconomic issues as their most important problem, but this changes in times of crisis.
- See the [Comparative Agendas Project](https://www.comparativeagendas.net/datasets_codebooks) for more information about this data and more datasets like it.



